---
title: "Data Science in R: Assignment 3"
author: "Steven Tran, Matthew David, Praveen Kenderla"
date: "February 18, 2018"
output:
    ioslides_presentation: null
    smaller: yes
    slidy_presentation: null
subtitle: R project 1
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
```

## Loading data sets into the R{.smaller}

```{r}
BP2010 <- read.csv("BP Apprehensions 2010.csv", header = T, stringsAsFactors = T)
PB2017 <- read.csv("PB Apprehensions 2017.csv", header = T, stringsAsFactors = T)
PBmonthly <- read.csv("monthly_sum.csv", header = T, stringsAsFactors = T)
rownames(PBmonthly) <- PBmonthly[,1]
```


## Converting rows into columns{.smaller} 
- columns indicate sectors
```{r}
a10 <- data.matrix(BP2010)
B10 <- matrix(NA, nrow = 13, ncol = 9)

c17 <- data.matrix(PB2017)
D17 <- matrix(NA, nrow = 13, ncol = 9)

for (i in 1:13){              # storing sectors as columns
  B10 [i, ] <- c(a10 [ , i])
  D17 [i, ] <- c(c17 [ , i])
}
```


## Barplot of months for year 2010{.smaller}

```{r fig.height = 4}
BPMonth_10 <- barplot(as.matrix(BP2010[1:9,2:13]),  beside = T,
                 las=2, ylab = "Apprehensions", xlab = "Months",
                 axisnames=TRUE,
                 main="2010 Border Patrol Apprehensions by Month",
                 border="blue")
```

## Barplot for months for year 2017{.smaller}

```{r fig.height = 4}
BPMonth_17 <- barplot(as.matrix(PB2017[1:9,2:13]),  beside = T,
                 las=2, ylab = "Apprehensions", xlab = "Months",
                 axisnames=TRUE,
                 main="2017 Border Patrol Apprehensions by Month",
                 border="red")
```

## Barplot for sectors for year 2010{.smaller}

``` {r fig.height = 4}
BPSector_10 <- barplot(B10[2:13, 1:9],  beside = T, names.arg = BP2010$Sector,
               las=2, axisnames=TRUE, xlab = "Sectors", ylab = "Apprehensions",
               main="2010 Border Patrol Apprehensions by Sector",col = rainbow (12))
legend ("topleft", inset = 0.02, title = "Months", colnames(BP2010 [2: 13]), 
        fill = rainbow (12), horiz = F, cex = 0.5)

```

## Barplot for sectors for year 2017{.smaller}

``` {r fig.height = 4}
BPSector_17 <- barplot(D17[2:13, 1:9],  beside = T, names.arg = BP2010$Sector,
               las=2, axisnames=TRUE, xlab = "Sectors", ylab = "Apprehensions",
               main="2017 Border Patrol Apprehensions by Sector", col = rainbow (12))
legend ("topright", inset = 0.02, title = "Months", colnames(BP2010 [2: 13]), 
        fill = rainbow (12), horiz = F, cex = 0.5)
```

## Apprehension data for different months for year 2010 and 2017 {.smaller}

```{r}
sideBySideMatrix <- function(month){
  matrix(c(BP2010[1:9,month], PB2017[1:9,month]), nrow = 2, byrow = TRUE)
}

sideBySideBarPlot <- function(month, monthString){
  barplot(sideBySideMatrix(month), names.arg = rownames(BP2010),
          las=2,
          axisnames=TRUE,
          beside=TRUE,
          col=c("blue", "red"),
          main = paste("Border Patrol Apprehensions in", monthString, sep=" "))
  legend("topleft", 
         c("2010", "2017"), 
         fill = c("blue", "red"))
}
```
### Function for creating side by side plots

## Plots of months and sector for year 2010 and 2017{.smaller}

``` {r fig.height = 8}
par(mfrow=c(4,3)); 
sideBySideBarPlot(2, "Oct"); sideBySideBarPlot(3, "Novr")
sideBySideBarPlot(4, "Dec"); sideBySideBarPlot(5, "Jan"); 
sideBySideBarPlot(6, "Feb"); sideBySideBarPlot(7, "March"); 
```

## Plotting the data in a form of grid {.smaller}

``` {r fig.height = 8}
par(mfrow=c(4,3));
sideBySideBarPlot(8, "April"); sideBySideBarPlot(9, "May"); 
sideBySideBarPlot(10, "June"); sideBySideBarPlot(11, "July"); 
sideBySideBarPlot(12, "August"); sideBySideBarPlot(13, "September")
par(mfrow=c(1,1))
```

## T-test{.smaller}

- For Tucson in 2010 and Rio Grande Valley in 2017, the sectors with the most apprehensions in those years

``` {r}
t2010 <- t(BP2010)
t2017 <- t(PB2017)
t.test(as.numeric(t2010[2:13, 8]), as.numeric(t2017[2:13, 6]))
```
### t-test results shows that maximum apprehensions in years 2010 and 2017 has not changed sig

## Altering table to see sum of three month periods{.smaller} 

``` {r}
m2010 <- subset(BP2010, select = -c(Sector))
m2010 <- rbind(m2010, colSums(m2010))
m2010 <- cbind(m2010, rowSums(m2010[1:3]))
m2010 <- cbind(m2010, rowSums(m2010[4:6]))
m2010 <- cbind(m2010, rowSums(m2010[7:9]))
m2010 <- cbind(m2010, rowSums(m2010[10:12]))
colnames(m2010) <- c(colnames(m2010)[1:12], "Oct-Dec", "Jan-Mar", "Apr-Jun", "Jul-Aug")

m2017 <- subset(PB2017, select = -c(Sector))
m2017 <- rbind(m2017, colSums(m2017))
m2017 <- cbind(m2017, rowSums(m2017[1:3]))
m2017 <- cbind(m2017, rowSums(m2017[4:6]))
m2017 <- cbind(m2017, rowSums(m2017[7:9]))
m2017 <- cbind(m2017, rowSums(m2017[10:12]))
colnames(m2017) <- c(colnames(m2017)[1:12], "Oct-Dec", "Jan-Mar", "Apr-Jun", "Jul-Aug") 
```

## t-test{.smaller}
- T-test for Jan-Mar period for 2010 and Oct-Dec period for 2017
``` {r}
t.test(m2010[1:9, 4:6], m2017[1:9, 1:3])
```
### Maximum sum of 3 months of apprehensions for both 2010 and 2017 are not significantly different

## Plotting time series data{.smaller}
class: small-code
- Continuous plot for months for both years

``` {r}
twolineplot <- function(){
  x <-  factor(2:13, labels = c("Oct", "Nov", "Dec", "Jan", "Feb", "Mar", "April",
                                "May","June","July","Aug","Sep"))
  t2010 <- as.numeric(PBmonthly[1,2:13])
  t2017 <- as.numeric(PBmonthly[8, 2:13])
  plot.default(t2010 ~ x, type="n", xlab = "Month", ylab = "Apprehensions", 
               main = "Apprehensions By Month", ylim = c(min(t2010),max(t2017)), xaxt = "n")
  axis(1, at = as.numeric(x), labels = levels(x))
  lines(t2010 ~ x, col="blue")
  lines(t2017 ~ x, col="red")
  legend("topleft", 
         c("2010", "2017"), 
         fill = c("blue", "red"))
}
```

## Plotting time series data (continued){.smaller}

```{r fig.height= 3.5, fig.width = 6}
twolineplot()
```

## Time Series data{.smaller}
``` {r}
tsplot <- function(){
ts1 <- data.matrix(subset(PBmonthly, select = -c(year)))
ts2 <- as.vector(rev(t(ts1)))
ts3 <- ts(ts2, start = c(2000,10), frequency = 12)
ts.plot(ts3, gpars=list(xlab="Year", ylab="Apprehensions", 
                        main = "Average Apprehension per Year", lty=c(1:3)))
## Adds the mean marks for each year
y <- as.vector(rev(rowMeans(PBmonthly)))
x <- 2000:2017
text(x+1, y-5000, labels = x, col = "blue", cex = 0.8)
segments(x, y, x+2, y, col = "red")
}
```
## Time Series Plot

```` {r fig.height = 5,fig.width = 8}
tsplot()
```

